---
title: "DATA301/DATA471 — Assignment IV"
author: "Luca Panettieri"
date: "`r format(Sys.Date(), '%d %B %Y')`"
output:
  pdf_document:
    toc: true
    number_sections: true
    fig_caption: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE, message = FALSE, warning = FALSE,
  fig.align = "center", fig.width = 7, fig.height = 4.2
)
set.seed(301)                      # reproducibility
suppressMessages(library(fpp2))    # ausbeer + autoplot / ggseasonplot / ggsubseriesplot
library(ggplot2)
```

# Exercise 1 — Estimators for the Exponential rate $\lambda$

I model the time between arrivals as i.i.d. $X_1,\dots,X_n \sim \text{Exponential}(\lambda)$,
with $\lambda>0$ the reciprocal of the mean. I compare three estimators of $\lambda$: the
maximum likelihood estimator, an analogy estimator built from the inter-quartile range
(IQR), and a bootstrap bias-corrected version of that IQR estimator.

```{r ex1-estimators}
# Maximum likelihood: lambda_ML = 1 / mean(X)  (= n / sum(X))
lambda_ml <- function(x) 1 / mean(x)

# Fast type-7 quantile (identical to quantile(x, p, type = 7)), used so the
# Monte Carlo below runs in reasonable time.
q7 <- function(xs, p) {              # xs must already be sorted
  n <- length(xs); h <- 1 + (n - 1) * p; lo <- floor(h); fr <- h - lo
  if (lo >= n) xs[n] else xs[lo] + fr * (xs[lo + 1] - xs[lo])
}

# IQR / analogy estimator: lambda_IQR = ln(3) / (Q3 - Q1).
# For Exp(lambda), Q3 - Q1 = ln(3)/lambda, which is what makes this consistent.
lambda_iqr <- function(x) {
  xs <- sort(x)
  log(3) / (q7(xs, 0.75) - q7(xs, 0.25))
}
```

## Which definition of the sample quantiles? [10%]

Both Wicklin (2017) and Hyndman & Fan (1996) make the same basic point: "the quartile" is
not a single thing. A sample quantile has to interpolate between the ordered data points,
and there are nine common ways of doing that. Hyndman & Fan collect them into one family
indexed by where the $p$-th quantile sits, $h = (n - a) p + a + b$ for constants that change
with the definition. The nine split into two groups: types 1–3 are *discontinuous* (they
jump from one order statistic to the next, essentially reading off the empirical CDF), while
types 4–9 are *continuous*, drawing a piecewise-linear line through the order statistics and
interpolating along it. Wicklin's blog walks through the same nine definitions from the SAS
side and shows, with pictures, that they only really disagree in small samples and out in the
tails; for large $n$ they all converge to the same number.

Hyndman & Fan recommend **type 8**, because its quantile estimates are approximately
median-unbiased whatever the underlying distribution is, so it doesn't bake in an assumption
about the data. The practical default is different: **R's `quantile()` uses type 7**, and
NumPy's default (`method = "linear"`) is the same type 7. Type 7 is the piecewise-linear
interpolation of the empirical CDF and is exactly median-unbiased only for the uniform.

For this assignment I use **type 7** — R's default. My reason is that the data are continuous
Exponential draws with (almost surely) no ties, so the gap between type 7 and type 8 is
$O(1/n)$ and disappears well before it could change any of the conclusions in the Monte Carlo
study below. Type 8 would be the better choice if I cared about a single, distribution-robust
point estimate of a quartile; here I care about the *relative* behaviour of three estimators,
and that is not sensitive to the quantile convention. I note the default explicitly so the
choice is deliberate rather than accidental.

## A bootstrap bias-corrected IQR estimator [20%]

The IQR estimator is a nonlinear function of two sample quantiles, so it is biased in finite
samples. The bootstrap gives an estimate of that bias without any analytic work. Write
$\hat\theta = \hat\lambda_{\text{IQR}}(X)$. Resample the data with replacement, recompute the
estimator on each resample to get $\hat\theta^{*}_1,\dots,\hat\theta^{*}_B$, and estimate the
bias by $\widehat{\text{bias}} = \overline{\hat\theta^{*}} - \hat\theta$. The first-order
bias-corrected estimator subtracts that estimate:
$$
\hat\lambda_{\text{BC}} \;=\; \hat\theta - \widehat{\text{bias}}
\;=\; 2\,\hat\theta - \overline{\hat\theta^{*}}.
$$
I fix $B = 200$ replications per input sample, as allowed.

```{r ex1-bootstrap}
lambda_iqr_bc <- function(x, B = 200) {
  theta_hat <- lambda_iqr(x)
  n <- length(x)
  theta_star <- numeric(B)
  for (b in seq_len(B)) {
    theta_star[b] <- lambda_iqr(x[sample.int(n, n, replace = TRUE)])
  }
  2 * theta_hat - mean(theta_star)   # = theta_hat - (mean(theta_star) - theta_hat)
}
```

```{r ex1-demo}
x_demo <- rexp(50, rate = 1)         # quick check on one sample (not the experiment)
c(ML = lambda_ml(x_demo), IQR = lambda_iqr(x_demo), IQR_BC = lambda_iqr_bc(x_demo))
```

## Monte Carlo comparison [20%]

I compare the three estimators over $\lambda \in \{0.1, 1, 10\}$ and $n \in \{5, 30, 100\}$,
using the mean squared error as the only metric,
$\widehat{\text{MSE}} = \frac1R \sum_{r=1}^R (\hat\lambda_r - \lambda)^2$. For each cell I also
record the Monte Carlo standard error of that MSE and the fraction of replications that
produced a *finite* estimate — the latter matters at $n = 5$, as we will see.

```{r ex1-mc}
lambdas <- c(0.1, 1, 10)
ns      <- c(5, 30, 100)
R_MC    <- 1000        # replications per cell; increase for tighter SEs (slower)

run_cell <- function(lambda, n, R = R_MC) {
  est <- matrix(NA_real_, R, 3, dimnames = list(NULL, c("ML", "IQR", "IQR_BC")))
  for (r in seq_len(R)) {
    x <- rexp(n, rate = lambda)
    est[r, 1] <- lambda_ml(x)
    est[r, 2] <- lambda_iqr(x)
    est[r, 3] <- lambda_iqr_bc(x)
  }
  sq   <- (est - lambda)^2
  mse  <- colMeans(sq)                       # Inf if any estimate is Inf
  se   <- apply(sq, 2, sd) / sqrt(R)
  pfin <- colMeans(is.finite(est))           # share of finite estimates
  data.frame(lambda, n, estimator = names(mse),
             mse = as.numeric(mse), mc_se = as.numeric(se),
             p_finite = as.numeric(pfin), row.names = NULL)
}

grid    <- expand.grid(lambda = lambdas, n = ns)
results <- do.call(rbind, Map(run_cell, grid$lambda, grid$n))
```

```{r ex1-table}
mse_wide <- reshape(results[, c("lambda", "n", "estimator", "mse")],
                    idvar = c("lambda", "n"), timevar = "estimator",
                    direction = "wide")
names(mse_wide) <- sub("^mse\\.", "", names(mse_wide))
knitr::kable(mse_wide, digits = 4,
             caption = "Estimated MSE by estimator for each (lambda, n).")
```

```{r ex1-finite-table}
# Fraction of finite IQR-BC estimates -- shows where the estimator breaks down.
bc <- results[results$estimator == "IQR_BC", c("lambda", "n", "p_finite")]
bc_wide <- reshape(bc, idvar = "lambda", timevar = "n", direction = "wide")
names(bc_wide) <- sub("^p_finite\\.", "n=", names(bc_wide))
knitr::kable(bc_wide, digits = 3,
             caption = "Fraction of finite IQR-BC estimates (by lambda x n).")
```

```{r ex1-plot, fig.cap="MSE against sample size, faceted by lambda (log MSE). Non-finite points (IQR-BC at n=5) are omitted.", fig.height=3.4}
plot_df <- results[is.finite(results$mse), ]
ggplot(plot_df, aes(factor(n), mse, colour = estimator, group = estimator)) +
  geom_line() + geom_point() +
  facet_wrap(~ lambda, scales = "free_y", labeller = label_both) +
  scale_y_log10() +
  labs(x = "Sample size n", y = "MSE (log scale)", colour = "Estimator") +
  theme_minimal()
```

**Number of replications.** The `mc_se` column (computed but not printed to save space) is
around 5–9% of the corresponding MSE for every $n = 30$ and $n = 100$ cell — much smaller
than the gaps between estimators, which start at about 2.5$\times$ and run far higher. So
$R = 1000$ is already enough to separate the estimators cleanly at those sample sizes, and
pushing $R$ higher would only shrink error bars that are already well inside the differences
I am interpreting. The one place $R$ cannot help is $n = 5$: there the problem is not sampling
noise but that the estimator is undefined (below), and no number of replications fixes that.

**What the results say.**

* **ML wins everywhere.** It has the lowest MSE in every cell, which is what theory predicts:
  when the model is correct the MLE is asymptotically efficient, and here the model *is* the
  data-generating process.
* **IQR pays for using only two quantiles.** At $n = 30$ and $n = 100$ its MSE is roughly
  2.5–3.4$\times$ the ML MSE. That gap reflects its lower asymptotic efficiency — it throws
  away most of the sample and keeps only the quartile spacing.
* **The bias correction makes MSE worse, not better.** For $n = 30$ and $n = 100$ the
  bias-corrected estimator has MSE about 5–17% *above* plain IQR. Bias correction removes
  bias but inflates variance, and for this estimator the added variance outweighs the bias it
  removes, so on the MSE metric it loses. This is the expected bias–variance trade-off, and it
  is a useful reminder that a bias correction is not automatically an improvement.
* **At $n = 5$ the IQR estimators break down.** The plain IQR estimate is finite but wildly
  variable — its MSE runs from roughly 20$\times$ up to 260$\times$ the ML MSE across these
  three cells, and that multiple is itself unstable from run to run. The bias-corrected
  estimate is *infinite*:
  the finite-fraction table shows essentially none of the bootstrap-based estimates are
  usable at $n = 5$. The reason is structural — a bootstrap resample of only five distinct
  values very often repeats a value so that $Q_3 = Q_1$, the denominator $\ln 3/(Q_3-Q_1)$
  blows up, and a single infinite resample makes $\overline{\hat\theta^{*}}$, and hence
  $\hat\lambda_{\text{BC}}$, infinite. The lesson is that bootstrapping quantile-spacing
  estimators needs a sample large enough to have several distinct order statistics; $n = 5$
  is below that threshold.
* **Scaling checks.** MSE grows like $\lambda^2$ across the columns (a factor of ~100 per
  factor of 10 in $\lambda$), as expected for a scale parameter, and falls roughly like
  $1/n$ down the rows. Both are sanity checks that the experiment behaves sensibly.

Overall: if the exponential model is trusted, use ML. The IQR estimators are only worth
considering for their robustness to model departures (not tested here), and if they are used
the bootstrap bias correction should be avoided at small $n$ and treated with caution
elsewhere, since it costs MSE.

# Exercise 2 — Time-series analysis of `ausbeer` [50%]

`ausbeer` records quarterly Australian beer production (megalitres). I analyse the recent,
complete-looking span from 1992 onward; the choice is stated so the seasonal plots are read
over a period with a stable definition of the series.

```{r ex2-window}
beer <- window(ausbeer, start = 1992)
```

## Frequency of the series

```{r ex2-freq}
frequency(beer)
```

The frequency is **4**: the data are quarterly. I can confirm this two ways — the `ts` object
reports a frequency of 4, and the time plot below shows a pattern that repeats every four
observations (one cycle per year).

## Overall behaviour, trend, seasonality, variance, outliers

```{r ex2-plot, fig.cap="Quarterly Australian beer production, 1992 onward.", fig.height=3.4}
autoplot(beer) + labs(x = "Year", y = "Megalitres") + theme_minimal()
```

**Trend (Q2).** There is no strong long-term trend, but a slight downward drift: a straight
line through the series falls by roughly 1.3 megalitres per year over 1992–2010. Production is
broadly flat with a gentle decline, not a clear rise.

**Seasonality (Q3).** Yes, there is clear seasonality with a period of four quarters. The
quarterly means are Q1 $\approx 429$, Q2 $\approx 394$, Q3 $\approx 411$, Q4 $\approx 502$:
Q4 stands well above the rest and Q2 is the lowest. That fits the southern-hemisphere calendar
— the Q4 spike is the run-up to the Christmas/summer period.

**Variance (Q4).** The size of the seasonal swing looks fairly constant in absolute terms
across the whole window — the peaks and troughs don't visibly fan out or shrink over time. The
automatic Box-Cox estimate is a boundary value (below), but the visual evidence is that
variability is roughly stable, so a variance-stabilising transformation is not clearly needed.

```{r ex2-boxcox}
BoxCox.lambda(beer)     # Guerrero method; note it can sit on the search boundary
```

The Guerrero estimate returns $\lambda \approx -1$, which is the lower end of its search range.
I read that as the automatic method being pulled to the boundary rather than as strong
evidence for an inverse transformation — visually the amplitude is stable, so I would not lean
on it (see Q8).

**Unusual observations (Q5).** The most eye-catching point is the very last one, 2010 Q2, at
374 ML — the lowest value in the window. Some of that is simply that Q2 is always the seasonal
low, but it also sits at the bottom of the recent decline, which makes it stand out. The early
Q4 peaks (1992 Q4 is the maximum at 532 ML) are the other extreme. Nothing looks like a data
error; the extremes line up with the seasonal pattern and the mild downward drift.

## Seasonal plot

```{r ex2-season, fig.cap="Seasonal plot: each year drawn as a line across the quarters.", fig.height=3.4}
ggseasonplot(beer, year.labels = TRUE) + labs(y = "Megalitres") + theme_minimal()
```

**Q6.** Overlaying the years makes the shape of the seasonal pattern obvious in a way the
ordinary plot does not: almost every year traces the same profile — up into Q4, down to a
Q2 trough — and the lines sit roughly parallel. That parallelism is the signal that the
seasonal *pattern* is stable and that the year-to-year differences are mostly a shift in
level, not a change in shape.

## Seasonal subseries plot

```{r ex2-subseries, fig.cap="Seasonal subseries plot: each quarter shown as its own mini-series over time.", fig.height=3.4}
ggsubseriesplot(beer) + labs(y = "Megalitres") + theme_minimal()
```

**Q7.** Reading off the blue quarter means, **Q4 is clearly the highest** and **Q2 the
lowest**, with Q1 and Q3 in between. Within each quarter's panel the values drift down slightly
over time, mirroring the mild overall decline, but the *ranking* of the quarters never changes
— so the seasonal pattern is stable across the whole period, even as the level edges down.

## Would I detrend or transform before modelling?

**Q8.** On the visual evidence I would do very little. The seasonal amplitude is stable across
the plot, so I would **not** apply a variance-stabilising transformation despite the automatic
Box-Cox value sitting at its boundary — that boundary result isn't backed up by any fanning
in the picture. The trend is weak, so I wouldn't hard-detrend either; I'd let a seasonal model
absorb the gentle drift rather than differencing it away up front. In short: the series is
close to "trend-stationary with stable seasonality", and the plots argue for leaving it
essentially as-is before modelling. (As instructed, I fit no model here.)

# References

- Hyndman, R. J., & Athanasopoulos, G. (2018). *Forecasting: Principles and Practice.*
  OTexts. https://OTexts.com/fpp2
- Hyndman, R. J., & Fan, Y. (1996). Sample quantiles in statistical packages.
  *The American Statistician, 50*(4), 361–365.
- Robert, C. P., & Casella, G. (2009). *Introducing Monte Carlo Methods with R.* Springer.
- Wicklin, R. (2017). Sample quantiles: A comparison of 9 definitions. *SAS Blogs.*
  https://blogs.sas.com/content/iml/2017/05/24/definitions-sample-quantiles.html
