---
title: "DATA301/DATA471 - Assignment IV"
author: "Luca Panettieri"
date: "`r format(Sys.Date(), '%d %B %Y')`"
output:
  pdf_document:
    toc: true
    number_sections: true
    fig_caption: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE, message = FALSE, warning = FALSE,
  fig.align = "center", fig.width = 7, fig.height = 4.2
)
set.seed(301)
suppressMessages(library(fpp2))
library(ggplot2)
```

# Exercise 1 - Estimators for the Exponential rate lambda

X_1, ..., X_n are independent and identically distributed as Exponential(lambda),
where lambda is greater than 0 and equals the reciprocal of the mean. I compare the
maximum likelihood estimator, the inter-quartile-range (IQR) analogy estimator, and a
bootstrap bias-corrected IQR estimator.

```{r ex1-estimators}
# Maximum likelihood: lambda_ML = 1 / mean(X)
lambda_ml <- function(x) 1 / mean(x)

# Fast type-7 quantile (identical to quantile(x, p, type = 7)), for speed.
q7 <- function(xs, p) {              # xs already sorted
  n <- length(xs); h <- 1 + (n - 1) * p; lo <- floor(h); fr <- h - lo
  if (lo >= n) xs[n] else xs[lo] + fr * (xs[lo + 1] - xs[lo])
}

# IQR estimator: lambda_IQR = ln(3) / (Q3 - Q1), since Q3 - Q1 = ln(3)/lambda.
lambda_iqr <- function(x) { xs <- sort(x); log(3) / (q7(xs, 0.75) - q7(xs, 0.25)) }
```

## Which definition of the sample quantiles? [10%]

Wicklin (2017) reviews nine definitions of the sample quantile: three discrete (which round
to order statistics) and six continuous (which interpolate linearly between them). They differ
mainly for small samples (n of 10 or fewer) and in the tails, and become indistinguishable by
about n = 100. Defaults vary by software (SAS uses type 2; R uses type 7). Wicklin does not
recommend a single type, but points to Hyndman and Fan (1996) for anyone who wants an unbiased
or distribution-free choice. Hyndman and Fan recommend type 8, which is approximately
median-unbiased for any distribution.

I use type 7, which is R's default (NumPy's "linear" method is the same). For continuous
Exponential data with no ties the difference between the types is of order 1/n and cannot
affect the comparison of estimators below, so the default is fine.

## A bootstrap bias-corrected IQR estimator [20%]

The IQR estimator is a nonlinear function of two quantiles, so it is biased in finite samples.
Write theta_hat for lambda_IQR(X). The bootstrap (Efron, 1979) estimates the bias as
(mean of theta_star) minus theta_hat, where theta_star_1, ..., theta_star_B are the estimator
recomputed on resamples of the data. The first-order bias-corrected estimator subtracts that
estimate of the bias:

    lambda_BC = theta_hat - (mean(theta_star) - theta_hat) = 2*theta_hat - mean(theta_star),

with B = 200 bootstrap resamples.

```{r ex1-bootstrap}
lambda_iqr_bc <- function(x, B = 200) {
  theta_hat  <- lambda_iqr(x)
  n <- length(x)
  theta_star <- replicate(B, lambda_iqr(x[sample.int(n, n, replace = TRUE)]))
  2 * theta_hat - mean(theta_star)
}
```

## Monte Carlo comparison [20%]

Over lambda in {0.1, 1, 10} and n in {5, 30, 100}, with R replications per cell, I draw a fresh
sample each replication and record the mean squared error (the sole metric). The MSE estimate
for a cell is the average over the R replications of (lambda_hat - lambda) squared. I also
record its Monte Carlo standard error.

```{r ex1-mc}
lambdas <- c(0.1, 1, 10); ns <- c(5, 30, 100)
R_MC <- 1000

run_cell <- function(lambda, n, R = R_MC) {
  est <- matrix(NA_real_, R, 3, dimnames = list(NULL, c("ML", "IQR", "IQR_BC")))
  for (r in seq_len(R)) {
    x <- rexp(n, rate = lambda)
    est[r, ] <- c(lambda_ml(x), lambda_iqr(x), lambda_iqr_bc(x))
  }
  sq <- (est - lambda)^2
  data.frame(lambda, n, estimator = colnames(est),
             mse = colMeans(sq), mc_se = apply(sq, 2, sd) / sqrt(R),
             row.names = NULL)
}

results <- do.call(rbind, Map(run_cell,
                              expand.grid(lambda = lambdas, n = ns)$lambda,
                              expand.grid(lambda = lambdas, n = ns)$n))
```

```{r ex1-table}
mse_wide <- reshape(results[, c("lambda", "n", "estimator", "mse")],
                    idvar = c("lambda", "n"), timevar = "estimator", direction = "wide")
names(mse_wide) <- sub("^mse\\.", "", names(mse_wide))
knitr::kable(mse_wide, digits = 4, caption = "Estimated MSE by estimator for each (lambda, n).")
```

```{r ex1-plot, fig.cap="MSE against sample size, faceted by lambda (log scale). Non-finite points (IQR-BC at n=5) omitted.", fig.height=3.2}
ggplot(results[is.finite(results$mse), ],
       aes(factor(n), mse, colour = estimator, group = estimator)) +
  geom_line() + geom_point() +
  facet_wrap(~ lambda, scales = "free_y", labeller = label_both) +
  scale_y_log10() +
  labs(x = "Sample size n", y = "MSE (log scale)", colour = "Estimator") +
  theme_minimal()
```

Replications. Following Robert and Casella (2009, section 3.2), a Monte Carlo average converges
at a rate of order 1 over the square root of R, and its error is judged from the estimated
standard error: the estimate plus or minus two standard errors gives an approximate 95 percent
band by the Central Limit Theorem (section 4.2 makes the same assessment by monitoring the
variability of the runs). Applying that criterion, the reported mc_se is under 10 percent of
its MSE for every cell with n of 30 or more, which is far smaller than the gaps between the
estimators, so R = 1000 is enough there. A larger R cannot help at n = 5, where the estimator
is undefined rather than merely noisy. This value of 1000 also matches the working default
(Nsim = 10^3) used throughout Robert and Casella.

Results. ML has the lowest MSE in every cell, which is expected because it is efficient when
the model is correct. The IQR estimator is about 2.4 to 3.4 times worse than ML at n of 30 or
more, since it uses only the quartiles. The bias correction slightly increases the MSE (it
removes bias but adds more variance), and this holds in every finite cell. At n = 5 the IQR
estimators fail: IQR is extremely variable, and IQR-BC is infinite, because bootstrap resamples
of five values often give Q3 equal to Q1 (a zero denominator). The MSE scales like lambda
squared across the columns and falls like 1 over n down the rows, as expected.

# Exercise 2 - Time-series analysis of ausbeer [50%]

This is approached as exploratory data analysis (per the course announcement): the plots below
extract the information in the data, and no model is fitted. The ausbeer series is quarterly
Australian beer production in megalitres; I use the span from 1992 onward.

```{r ex2-window}
beer <- window(ausbeer, start = 1992)
```

## Frequency

```{r ex2-freq}
frequency(beer)
```

The frequency is 4 (quarterly); the plot also repeats every four observations.

## Behaviour, trend, seasonality, variance, outliers

```{r ex2-plot, fig.cap="Quarterly Australian beer production, 1992 onward.", fig.height=3.2}
autoplot(beer) + labs(x = "Year", y = "Megalitres") + theme_minimal()
```

- Trend: no strong trend, just a slight downward drift of about 1.3 megalitres per year.
- Seasonality: clear and periodic (a fixed four-quarter cycle, not merely cyclic). The quarter
  means are Q1 429, Q2 394, Q3 411, Q4 502, so Q4 is highest and Q2 is lowest.
- Variance: the seasonal amplitude is stable over time, so no transformation is needed.
- Unusual points: 2010 Q2 (374 megalitres) is the lowest and 1992 Q4 (532 megalitres) the
  highest; both are consistent with the seasonal pattern rather than being data errors.

## Seasonal plot

```{r ex2-season, fig.cap="Seasonal plot.", fig.height=3.2}
ggseasonplot(beer, year.labels = TRUE) + labs(y = "Megalitres") + theme_minimal()
```

Every year follows the same profile (a peak in Q4 and a trough in Q2) with roughly parallel
lines, so the seasonal shape is stable and the years differ mainly in their overall level.

## Seasonal subseries plot

```{r ex2-subseries, fig.cap="Seasonal subseries plot.", fig.height=3.2}
ggsubseriesplot(beer) + labs(y = "Megalitres") + theme_minimal()
```

Q4 is the highest quarter and Q2 the lowest (the blue quarter means). Each quarter drifts down
slightly over time, but the ranking never changes, so the seasonal pattern is stable.

## Detrend or transform before modelling?

No. The variance is stable (so no transformation) and the trend is weak; the only substantial
departure from stationarity is the seasonality, so if anything were done before modelling it
would be a single seasonal difference. This is a visual argument only, with no model fitted.

# References

- Efron, B. (1979). Bootstrap methods: Another look at the jackknife. The Annals of
  Statistics, 7(1), 1-26.
- Hyndman, R. J., and Athanasopoulos, G. (2018). Forecasting: Principles and Practice.
  OTexts. https://OTexts.com/fpp2
- Hyndman, R. J., and Fan, Y. (1996). Sample quantiles in statistical packages.
  The American Statistician, 50(4), 361-365.
- Robert, C. P., and Casella, G. (2009). Introducing Monte Carlo Methods with R. Springer.
- Wicklin, R. (2017). Sample quantiles: A comparison of 9 definitions. SAS Blogs.
  https://blogs.sas.com/content/iml/2017/05/24/definitions-sample-quantiles.html
