---
title: "DATA301/DATA471 — Assignment IV"
author: "Luca Panettieri"
date: "`r format(Sys.Date(), '%d %B %Y')`"
output:
  pdf_document:
    toc: true
    number_sections: true
    fig_caption: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.align = "center",
  fig.width = 7,
  fig.height = 4.5
)

# Reproducibility
set.seed(301)

# Packages
# install.packages("fpp2")  # run once if not installed
library(fpp2)      # ausbeer data + autoplot / ggseasonplot / ggsubseriesplot
library(ggplot2)
```

# Exercise 1 — Estimators for the Exponential rate $\lambda$

We model the time between successive customer arrivals as i.i.d.
$X_1,\dots,X_n \sim \text{Exponential}(\lambda)$, where $\lambda>0$ is the
reciprocal of the mean. We compare three estimators of $\lambda$:

- the **maximum likelihood** estimator,
- an **analogy** estimator based on the inter-quartile range (IQR), and
- a **bootstrap bias-corrected** version of the IQR estimator.

```{r ex1-estimators}
# Maximum likelihood estimator: lambda_ML = 1 / mean(X) = n / sum(X)
lambda_ml <- function(x) {
  1 / mean(x)
}

# IQR / analogy estimator: lambda_IQR = ln(3) / (Q3(X) - Q1(X))
# For Exp(lambda): Q3 - Q1 = ln(3)/lambda, hence the estimator below.
# `qtype` lets us switch quantile definition (see the [10%] discussion).
lambda_iqr <- function(x, qtype = 7) {
  q <- quantile(x, probs = c(0.25, 0.75), type = qtype, names = FALSE)
  log(3) / (q[2] - q[1])
}
```

## Order statistics: which quantile definition? [10%]

<!--
TODO (write-up, ~1/2 page):
- Summarise Wicklin (2017): there are 9 sample-quantile "types" (definitions),
  differing in how they interpolate between order statistics.
- Note Hyndman & Fan (1996) framework and their recommendation.
- State the DEFAULT: R's quantile() uses type 7; NumPy's default is also
  equivalent to type 7.
- Say which type you use and WHY (e.g. type 8: approximately median-unbiased and
  distribution-independent, as recommended by Hyndman & Fan).
- Whatever you choose, set QTYPE below so the whole assignment stays consistent.
-->

```{r ex1-qtype}
# Quantile definition used throughout this assignment.
# Change to 8 if you argue for the Hyndman & Fan recommendation.
QTYPE <- 7

# Show the default your R version reports (context for the discussion):
# ?quantile documents type = 7 as the default.
```

## Bootstrap bias-corrected IQR estimator [20%]

<!--
TODO (write-up, short):
- The IQR estimator is a nonlinear function of the sample quantiles, so it is
  biased in general.
- First-order bootstrap bias correction: given theta_hat = lambda_iqr(x),
  the bootstrap estimates the bias as mean(theta_star) - theta_hat, so the
  bias-corrected estimator subtracts that:
      lambda_BC = theta_hat - (mean(theta_star) - theta_hat)
                = 2*theta_hat - mean(theta_star).
- We fix B = 200 bootstrap replications per sample, as permitted.
-->

```{r ex1-bootstrap}
# First-order bootstrap bias-corrected IQR estimator.
# B = 200 bootstrap resamples per input sample.
lambda_iqr_bc <- function(x, B = 200, qtype = QTYPE) {
  theta_hat <- lambda_iqr(x, qtype = qtype)

  n <- length(x)
  theta_star <- replicate(B, {
    xb <- sample(x, size = n, replace = TRUE)
    lambda_iqr(xb, qtype = qtype)
  })

  # bias_hat = mean(theta_star) - theta_hat
  # lambda_BC = theta_hat - bias_hat = 2*theta_hat - mean(theta_star)
  2 * theta_hat - mean(theta_star)
}
```

```{r ex1-bootstrap-demo}
# Quick sanity check on a single sample (not part of the marked experiment).
x_demo <- rexp(50, rate = 1)
c(
  ML     = lambda_ml(x_demo),
  IQR    = lambda_iqr(x_demo, qtype = QTYPE),
  IQR_BC = lambda_iqr_bc(x_demo)
)
```

## Monte Carlo comparison [20%]

We compare the three estimators by Monte Carlo over the parameter grid
$\lambda \in \{0.1, 1, 10\}$ and $n \in \{5, 30, 100\}$, using the
**mean squared error** as the sole metric:
$\widehat{\text{MSE}} = \frac{1}{R}\sum_{r=1}^{R}(\hat\lambda_r - \lambda)^2$.

```{r ex1-monte-carlo}
lambdas <- c(0.1, 1, 10)
ns      <- c(5, 30, 100)

# Number of Monte Carlo replications per (lambda, n) cell.
# Justify this choice in the write-up using the MC standard error reported below.
R_MC <- 2000

run_cell <- function(lambda, n, R = R_MC) {
  # Matrix of estimates: rows = replications, cols = estimators
  est <- matrix(NA_real_, nrow = R, ncol = 3,
                dimnames = list(NULL, c("ML", "IQR", "IQR_BC")))

  for (r in seq_len(R)) {
    x <- rexp(n, rate = lambda)
    est[r, "ML"]     <- lambda_ml(x)
    est[r, "IQR"]    <- lambda_iqr(x, qtype = QTYPE)
    est[r, "IQR_BC"] <- lambda_iqr_bc(x)
  }

  sq_err <- (est - lambda)^2                 # squared errors, per estimator
  mse    <- colMeans(sq_err)                 # MSE estimate
  mc_se  <- apply(sq_err, 2, sd) / sqrt(R)   # Monte Carlo SE of each MSE

  data.frame(
    lambda    = lambda,
    n         = n,
    estimator = names(mse),
    mse       = as.numeric(mse),
    mc_se     = as.numeric(mc_se),
    row.names = NULL
  )
}

# Run the full grid (this is the slow chunk: 3 x 3 x R_MC x B bootstraps).
grid    <- expand.grid(lambda = lambdas, n = ns)
results <- do.call(
  rbind,
  Map(run_cell, grid$lambda, grid$n)
)

results
```

```{r ex1-results-table}
# Wide MSE table (rows = lambda x n, cols = estimator) for the report.
mse_wide <- reshape(
  results[, c("lambda", "n", "estimator", "mse")],
  idvar     = c("lambda", "n"),
  timevar   = "estimator",
  direction = "wide"
)
names(mse_wide) <- sub("^mse\\.", "", names(mse_wide))
knitr::kable(
  mse_wide,
  digits  = 4,
  caption = "Estimated MSE by estimator, for each (lambda, n)."
)
```

```{r ex1-results-plot, fig.cap="MSE by sample size, faceted by lambda (log scale on MSE)."}
ggplot(results, aes(x = factor(n), y = mse,
                    colour = estimator, group = estimator)) +
  geom_line() +
  geom_point() +
  facet_wrap(~ lambda, scales = "free_y",
             labeller = label_both) +
  scale_y_log10() +
  labs(x = "Sample size n", y = "MSE (log scale)", colour = "Estimator") +
  theme_minimal()
```

<!--
TODO (write-up):
- Justify R_MC: point to the mc_se column — argue R is large enough that the
  differences in MSE between estimators are not Monte Carlo noise (the gaps are
  large relative to the reported standard errors).
- Interpret: which estimator wins where. Expect ML to dominate (model is correct),
  IQR to be worst, and the bias correction to help IQR most at small n.
-->

# Exercise 2 — Time-series analysis of `ausbeer` [50%]

The `ausbeer` series (from `fpp2`) records quarterly beer production in
Australia. This is a plots-and-interpretation exercise — **no model is fitted.**

```{r ex2-load}
# Full series runs to 2010 Q2; the last two quarters are incomplete-year context.
# Window to a clean, recent span for the seasonal discussion (state your choice).
beer <- window(ausbeer, start = 1992)
```

## Frequency of the series

```{r ex2-frequency}
frequency(beer)
```

<!--
TODO: frequency = 4 (quarterly). Explain you can also read this off the plot from
the repeating 4-quarter cycle.
-->

## Overall behaviour, trend, seasonality, variance, outliers

```{r ex2-autoplot, fig.cap="Quarterly Australian beer production."}
autoplot(beer) +
  labs(x = "Year", y = "Megalitres") +
  theme_minimal()
```

<!--
TODO — answer from the plot above:
- Q2 Trend: describe the long-term movement (roughly flat / slight decline over
  this window).
- Q3 Seasonality: yes; describe the regular within-year peaks and troughs.
- Q4 Variance: is the size of the seasonal swing roughly constant? If it changes,
  note a transformation (log / Box-Cox) might help. BoxCox.lambda below supports this.
- Q5 Unusual observations: point to any obvious spikes/dips and explain.
-->

```{r ex2-boxcox}
# A transformation (not a model) — supports the variance / Q8 discussion.
BoxCox.lambda(beer)
```

## Seasonal plot

```{r ex2-seasonplot, fig.cap="Seasonal plot: each year overlaid by quarter."}
ggseasonplot(beer, year.labels = TRUE) +
  labs(y = "Megalitres") +
  theme_minimal()
```

<!--
TODO (Q6): what does overlaying years reveal that the ordinary plot does not?
Comment on the consistent quarter ranking and any years that break the pattern.
-->

## Seasonal subseries plot

```{r ex2-subseries, fig.cap="Seasonal subseries plot: each quarter over time."}
ggsubseriesplot(beer) +
  labs(y = "Megalitres") +
  theme_minimal()
```

<!--
TODO (Q7): which quarters have the highest / lowest typical values (blue mean line)?
Is the seasonal pattern stable across years? (Expect Q4 highest, Q1/Q2 lower.)
-->

## Detrend or transform before modelling?

<!--
TODO (Q8): give a VISUAL argument (referencing the plots above + the Box-Cox lambda)
for whether you would detrend and/or transform the series before modelling.
Do NOT fit a model.
-->

# References

- Hyndman, R. J., & Athanasopoulos, G. (2018). *Forecasting: Principles and
  Practice*. OTexts. https://OTexts.com/fpp2
- Hyndman, R. J., & Fan, Y. (1996). Sample quantiles in statistical packages.
  *The American Statistician, 50*(4), 361–365.
- Robert, C. P., & Casella, G. (2009). *Introducing Monte Carlo Methods with R.*
  Springer.
- Wicklin, R. (2017). Sample quantiles: A comparison of 9 definitions. *SAS Blogs.*
  https://blogs.sas.com/content/iml/2017/05/24/definitions-sample-quantiles.html
