---
title: "DATA301/DATA471 — Assignment IV"
author: "Luca Panettieri"
date: "`r format(Sys.Date(), '%d %B %Y')`"
output:
  pdf_document:
    toc: true
    number_sections: true
    fig_caption: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE, message = FALSE, warning = FALSE,
  fig.align = "center", fig.width = 7, fig.height = 4.2
)
set.seed(301)
suppressMessages(library(fpp2))
library(ggplot2)
```

# Exercise 1 — Estimators for the Exponential rate $\lambda$

$X_1,\dots,X_n \sim \text{Exponential}(\lambda)$ i.i.d., with $\lambda>0$ the reciprocal of the
mean. I compare the maximum likelihood estimator, the inter-quartile-range (IQR) analogy
estimator, and a bootstrap bias-corrected IQR estimator.

```{r ex1-estimators}
# Maximum likelihood: lambda_ML = 1 / mean(X)
lambda_ml <- function(x) 1 / mean(x)

# Fast type-7 quantile (identical to quantile(x, p, type = 7)), for speed.
q7 <- function(xs, p) {              # xs already sorted
  n <- length(xs); h <- 1 + (n - 1) * p; lo <- floor(h); fr <- h - lo
  if (lo >= n) xs[n] else xs[lo] + fr * (xs[lo + 1] - xs[lo])
}

# IQR estimator: lambda_IQR = ln(3) / (Q3 - Q1), since Q3 - Q1 = ln(3)/lambda.
lambda_iqr <- function(x) { xs <- sort(x); log(3) / (q7(xs, 0.75) - q7(xs, 0.25)) }
```

## Which definition of the sample quantiles? [10%]

Wicklin (2017) reviews nine definitions of the sample quantile: three *discrete* (rounding to
order statistics) and six *continuous* (linear interpolation between them). They differ mainly
for small samples ($n \le 10$) and in the tails, and become indistinguishable by $n \approx
100$. Defaults vary by software (SAS uses type 2; R uses type 7); Wicklin does not prescribe a
type, but points to Hyndman & Fan (1996) for anyone wanting an unbiased or distribution-free
choice. Hyndman & Fan recommend **type 8**, which is approximately median-unbiased for any
distribution.

I use **type 7** — R's default (NumPy's `"linear"` method is equivalent). For continuous
Exponential data with no ties the difference between types is $O(1/n)$ and cannot affect the
comparison of estimators below, so the default is adequate.

## A bootstrap bias-corrected IQR estimator [20%]

$\hat\lambda_{\text{IQR}}$ is a nonlinear function of two quantiles, so it is biased in finite
samples. With $\hat\theta = \hat\lambda_{\text{IQR}}(X)$, the bootstrap (Efron, 1979) estimates
the bias as $\overline{\hat\theta^{*}} - \hat\theta$ from resampled estimates
$\hat\theta^{*}_1,\dots,\hat\theta^{*}_B$, giving the first-order corrected estimator
$$
\hat\lambda_{\text{BC}} = \hat\theta - (\overline{\hat\theta^{*}} - \hat\theta)
= 2\hat\theta - \overline{\hat\theta^{*}}, \qquad B = 200.
$$

```{r ex1-bootstrap}
lambda_iqr_bc <- function(x, B = 200) {
  theta_hat  <- lambda_iqr(x)
  n <- length(x)
  theta_star <- replicate(B, lambda_iqr(x[sample.int(n, n, replace = TRUE)]))
  2 * theta_hat - mean(theta_star)
}
```

## Monte Carlo comparison [20%]

Over $\lambda \in \{0.1, 1, 10\}$ and $n \in \{5, 30, 100\}$, with $R$ replications per cell, I
draw a fresh sample each replication and record the mean squared error (the sole metric),
$\widehat{\text{MSE}} = \frac1R \sum_{r=1}^R (\hat\lambda_r - \lambda)^2$, plus its Monte Carlo
standard error.

```{r ex1-mc}
lambdas <- c(0.1, 1, 10); ns <- c(5, 30, 100)
R_MC <- 1000

run_cell <- function(lambda, n, R = R_MC) {
  est <- matrix(NA_real_, R, 3, dimnames = list(NULL, c("ML", "IQR", "IQR_BC")))
  for (r in seq_len(R)) {
    x <- rexp(n, rate = lambda)
    est[r, ] <- c(lambda_ml(x), lambda_iqr(x), lambda_iqr_bc(x))
  }
  sq <- (est - lambda)^2
  data.frame(lambda, n, estimator = colnames(est),
             mse = colMeans(sq), mc_se = apply(sq, 2, sd) / sqrt(R),
             row.names = NULL)
}

results <- do.call(rbind, Map(run_cell,
                              expand.grid(lambda = lambdas, n = ns)$lambda,
                              expand.grid(lambda = lambdas, n = ns)$n))
```

```{r ex1-table}
mse_wide <- reshape(results[, c("lambda", "n", "estimator", "mse")],
                    idvar = c("lambda", "n"), timevar = "estimator", direction = "wide")
names(mse_wide) <- sub("^mse\\.", "", names(mse_wide))
knitr::kable(mse_wide, digits = 4, caption = "Estimated MSE by estimator for each (lambda, n).")
```

```{r ex1-plot, fig.cap="MSE against sample size, faceted by lambda (log scale). Non-finite points (IQR-BC at n=5) omitted.", fig.height=3.2}
ggplot(results[is.finite(results$mse), ],
       aes(factor(n), mse, colour = estimator, group = estimator)) +
  geom_line() + geom_point() +
  facet_wrap(~ lambda, scales = "free_y", labeller = label_both) +
  scale_y_log10() +
  labs(x = "Sample size n", y = "MSE (log scale)", colour = "Estimator") +
  theme_minimal()
```

**Replications.** Following Robert & Casella (2009, §3.2), a Monte Carlo average converges at
rate $O(1/\sqrt{R})$ and its error is judged from the estimated standard error $\sqrt{v_R}$,
with $\pm 2\sqrt{v_R}$ giving an approximate 95% band by the Central Limit Theorem (§4.2 makes
the same assessment by monitoring the variability of the runs). Applying that criterion, the
reported `mc_se` (an estimate of $\sqrt{v_R}$ for each MSE) is under 10% of its MSE for every
$n \ge 30$ cell — far smaller than the gaps between estimators — so $R = 1000$ is sufficient
there. Larger $R$ cannot help at $n = 5$, where the estimator is undefined rather than merely
noisy.

**Results.** ML has the lowest MSE in every cell (it is efficient when the model is correct).
The IQR estimator is about 2.4–3.4$\times$ worse than ML at $n \ge 30$, since it uses only the
quartiles. The bias correction slightly *increases* MSE (it removes bias but adds more
variance). At $n = 5$ the IQR estimators fail: IQR is extremely variable, and IQR-BC is
infinite because bootstrap resamples of five values often give $Q_3 = Q_1$ (zero denominator).
MSE scales like $\lambda^2$ and falls like $1/n$, as expected.

# Exercise 2 — Time-series analysis of `ausbeer` [50%]

Approached as exploratory data analysis (per the course announcement): the plots below extract
the information in the data; no model is fitted. `ausbeer` is quarterly Australian beer
production (megalitres); I use the span from 1992 onward.

```{r ex2-window}
beer <- window(ausbeer, start = 1992)
```

## Frequency

```{r ex2-freq}
frequency(beer)
```

Frequency 4 (quarterly); the plot also repeats every four observations.

## Behaviour, trend, seasonality, variance, outliers

```{r ex2-plot, fig.cap="Quarterly Australian beer production, 1992 onward.", fig.height=3.2}
autoplot(beer) + labs(x = "Year", y = "Megalitres") + theme_minimal()
```

- **Trend:** no strong trend — a slight downward drift (about $-1.3$ ML/year).
- **Seasonality:** clear and *periodic* (fixed four-quarter cycle, not merely cyclic). Quarter
  means: Q1 429, Q2 394, Q3 411, Q4 502 — Q4 highest, Q2 lowest.
- **Variance:** the seasonal amplitude is stable over time, so no transformation is needed.
- **Unusual points:** 2010 Q2 (374 ML) is the lowest and 1992 Q4 (532 ML) the highest; both
  are consistent with the seasonal pattern, not data errors.

## Seasonal plot

```{r ex2-season, fig.cap="Seasonal plot.", fig.height=3.2}
ggseasonplot(beer, year.labels = TRUE) + labs(y = "Megalitres") + theme_minimal()
```

Every year follows the same profile (peak in Q4, trough in Q2) with roughly parallel lines, so
the seasonal shape is stable and years differ mainly in level.

## Seasonal subseries plot

```{r ex2-subseries, fig.cap="Seasonal subseries plot.", fig.height=3.2}
ggsubseriesplot(beer) + labs(y = "Megalitres") + theme_minimal()
```

Q4 is highest and Q2 lowest (blue means); each quarter drifts down slightly over time but the
ranking never changes, so the seasonal pattern is stable.

## Detrend or transform before modelling?

No. The variance is stable (no transformation) and the trend is weak; the only substantial
non-stationarity is the seasonality, so if anything were done before modelling it would be a
single seasonal difference. Visual argument only — no model fitted.

# References

- Efron, B. (1979). Bootstrap methods: Another look at the jackknife. *The Annals of
  Statistics, 7*(1), 1–26.
- Hyndman, R. J., & Athanasopoulos, G. (2018). *Forecasting: Principles and Practice.*
  OTexts. https://OTexts.com/fpp2
- Hyndman, R. J., & Fan, Y. (1996). Sample quantiles in statistical packages.
  *The American Statistician, 50*(4), 361–365.
- Robert, C. P., & Casella, G. (2009). *Introducing Monte Carlo Methods with R.* Springer.
- Wicklin, R. (2017). Sample quantiles: A comparison of 9 definitions. *SAS Blogs.*
  https://blogs.sas.com/content/iml/2017/05/24/definitions-sample-quantiles.html
